Typing the Outputs

### 2015-08-13 sematics/boundary/boundary.tex
working on values and computations section

We implement first order recursive functions by introducing a new typing judgement \mbox{$\colfun f {A\to B} w$},
which says that $f$ is a function from $A$ to $B$ at world $w$.  
We use a different typing judgement here in order to keep functions and other values distinct.
We also allow let statements to work for second stage functions 
(for reasons explained later, first stage function lets can be safely eliminated \emph{a priori} by substitution).
Note how the only way to derive a function type is to directly declare a function, or to assume it.
\begin{framed}
\noindent\textbf{First-Order Recursive Functions:}
\begin{mathpar}
\infertypeswor [\to I]		{\Gamma \vdash \colfun {\fix fx{e_1}} {A \to B} w}	{\ty [,\colfun f {A \to B} w,\col x A] {e_1} B} 							\and
%\infertypeswor [fix]		{\ty {\fix fxe} {A \to B}}							{\ty [,\col f {A \to B},\col x A] e B} 										\and
\infertypeswor [\to E]		{\ty {\app {e_1}{e_2}} {B}}							{\Gamma \vdash \colfun {e_1} {A \to B} w & \ty {e_2} A} 					\and
\infertypeswor [\fut E]		{\typestwo {\letin x {e_1} {e_2}} B}				{\Gamma \vdash \colfun {e_1} {A \to B} \bbtwo 
																				&\typestwo [\Gamma,\colfun x {A \to B} \bbtwo] {e_2} B } 					\and
\end{mathpar}
\end{framed}

@@

...

>>>>

We implement first order recursive functions by introducing a new typing judgement \mbox{$\colfun f {A\to B} w$},
which says that $f$ is a function from $A$ to $B$ at world $w$.  
We use a different typing judgement here in order to keep functions and other values distinct.
That is, the $\to$ connective only ever appears to the right of ``$:>$'', and never to the right of a ``$:$".
We also allow let statements to work for second stage functions 
(for reasons explained later, first stage function lets can be safely eliminated \emph{a priori} by substitution).
Note how the only way to derive a function type is to directly declare a function, or to assume it.
\begin{framed}
\noindent\textbf{First-Order Recursive Functions:}
\begin{mathpar}
\infertypeswor [\to I]		{\Gamma \vdash \colfun {\fix fx{e_1}} {A \to B} w}	{\ty [,\colfun f {A \to B} w,\col x A] {e_1} B} 							\and
%\infertypeswor [fix]		{\ty {\fix fxe} {A \to B}}							{\ty [,\col f {A \to B},\col x A] e B} 										\and
\infertypeswor [\to E]		{\ty {\app {e_1}{e_2}} {B}}							{\Gamma \vdash \colfun {e_1} {A \to B} w & \ty {e_2} A} 					\and
\infertypeswor [\fut E]		{\typestwo {\letin f {e_1} {e_2}} B}				{\Gamma \vdash \colfun {e_1} {A \to B} \bbtwo 
																				&\typestwo [\Gamma,\colfun f {A \to B} \bbtwo] {e_2} B } 					\and
\end{mathpar}
\end{framed}

@@

It turns out that, for the purposes of splitting, it will be handy to differentiate terms into two classes: 
\emph{values} which are fully reduced, and \emph{computations} which may have some reduction remaining.
Importantly, we want to \emph{explicitly note} when subterms of a computation are actually values,
and so we introduce a new piece of syntax to mark just that inclusion: the underline constructor, written $\exv v$
where $v$ is a term (more precisely, a value).  
Note that the type of the underline is just the type of it's contents, which avoids interaction with the type system.

Because staging and the value/computation distinction are both related to reduction, they will interact non-trivially.
For this reason, we'll first devise a  system for a monostage language, before then considering a staged language.

\subsection{Monostage System}

Priori work has observed the difference between values and computations
via a syntactic distinction (i.e. saying there are two kinds of terms, value-terms and computation-terms).
We instead observe the difference judgementally: there's only one kind of term, 
but there are two judgements ($\valsym$ and $\compsym$) which classify terms as one or the other.

For instance, the aforementioned underline constructor always yields a computation,
but its contents are a value:

\begin{framed}
\noindent\textbf{Underlines:}
\begin{mathpar}
\infer {\exv v \comp} {v \val} \and
\infer {\exv v : A} {v : A} 
\end{mathpar}
\end{framed}

Int, bool, and unit base terms are always values.  
Intro forms, (tuples, injections, and rolls) are values so long as all their subterms are.

\begin{framed}
\noindent\textbf{Standard Values:}
\begin{mathpar}
\infer {\Gamma \vdash \tup{} \val}						{\cdot} 											\and
\infer {\Gamma \vdash i \val}							{\cdot}												\and
\infer {\Gamma \vdash b \val}							{\cdot}												\and
\infer {\Gamma \vdash \tup{e_1,e_2} \val}				{\Gamma \vdash e_1 \val & \Gamma \vdash e_2 \val} 	\and
\infer {\Gamma \vdash \inl e \val}						{\Gamma \vdash e \val} 								\and
\infer {\Gamma \vdash \inr e \val}						{\Gamma \vdash e \val} 								\and
\infer {\Gamma \vdash \roll e \val}						{\Gamma \vdash e \val} 										
\end{mathpar}
\end{framed}

Inversely, intro forms are also computations whenever all of the subterms are.
The elim forms, lets, and ifs are always computations.
Note that because we're studying a call-by-value language, all variables are introduced as values.

\begin{framed}
\noindent\textbf{Standard Computations:}
\begin{mathpar}
\infer {\Gamma \vdash \tup{e_1,e_2} \comp}				{\Gamma \vdash e_1 \comp & \Gamma \vdash e_2 \comp} 	\and
\infer {\Gamma \vdash \inl e \comp}						{\Gamma \vdash e \comp} 								\and
\infer {\Gamma \vdash \inr e \comp}						{\Gamma \vdash e \comp} 								\and
\infer {\Gamma \vdash \roll e \comp}					{\Gamma \vdash e \comp} 										
\end{mathpar}
\end{framed}
